OpenMotion: An open source library for attitude estimation.

### 2016-05-13 Article/prez_library.tex
modif article journal smartphone

Voici les 2 reference frame utiliser pour la creation de la librairie

@@

\item The object-fixed reference frame $\{c\}$ is a moving and rotating coordinate frame that is fixed to the smartphone device. 
\end{itemize}

The library admit admit that all the components of the IMU belong to the same reference frame $\{c\}$ (ref figure~\ref{Schema_situation}). It is a cyclopean approximation \cite{ouarti2008multimodal} .

@@

\texttt{OpenMotion} est une librairie open source qui permet a l'utilisateur d'estimer l'orientation du telephone (representer. En utilisant uniquement les donnee de l'IMU integre au telephone. La librairie ne supporte pas encore les translations, nous nous sommes concentre sur l'attitude.

>>>>

\texttt{OpenMotion} is a open source library mainly aimed for real-time attitude estimation. Our approach is based on data fusion technics using only the output of a IMU (composed by a 3D gyroscope,  3D accelerometer and a 3D magnetometer). Among several methods proposed by the library, the user has the possibilities to choose which methods fits the best to his expectation according to its performance. Here the 2 reference frame to remember.

@@

\item The object-fixed reference frame $\{c\}$ corresponding to the IMU device. It is a moving and rotating coordinate frame. 
\end{itemize}

For simplicity reasons, the library admit that all the components of the IMU belong to the same reference frame $\{c\}$ (ref figure \ref{Schema_situation}). It is a cyclopean approximation \cite{ouarti2008multimodal}. Moreover, the user has to interface the sends of data from the device to the library because the library does not take care of it. \texttt{OpenMotion} provide an estimation of the orientation of $\{c\}$ relative to $\{a\}$. The choice of the output type (quaternion, rotation matrix, euler angle) is according to the user needs.

@@

%
%\begin{lstlisting}[caption=Exemple of utilisation in C++,label=code]
%// om_exemple.cpp
%#include <om.h>
%
%using namespace om;
%
%void main(int argc, char** argv){
%
%	// 
%	SensorFusionManager manager;
%
%	//if calibration is necessary
%	manager.calibration();
%	
%	// while IMU send data
%	while( isData ){
%
%		// set all IMU data at time t
%		manager.setGyroscopeData(gyro_x,gyro_y,gyro_z);	
%		manager.setAccelerometerData(acc_x,acc_y,acc_z);
%		manager.setMagnetometerData(mag_x,mag_y,mag_z);
%		
%		// get different kind of output according to user's choices
%		
%		// quaternion
%		Quaternion q_est = manager.getAttitude<Quaternion>(); 
%		
%		//  rotation matrix
%		Matrix R = manager.getAttitude<Matrix>(); 
%		
%		//  axis angle
%		AxisAngle axisAngle = manager.getAttitude<Matrix>(); 
%		...
%	}
%}
%
%\end{lstlisting}

We are also working on the establishment of a dynamic sensor calibration process on all component of the IMU. We insist on the dynamic aspect. Indeed, firstly, it allows  the library to be adaptable to any kind of IMU, but also to improve the performance without requesting some effort from the user.

### 2016-05-13 Article/state_of_the_art.tex
modif article journal smartphone

\subsection{Kinematic and quaternion representation}

@@

For attitude representation, the quaternion must satisfies a single constraint given by  $|| \mathbi{q} || = 1$.The quaternion kinematics equation in discrete-time is given by :

\begin{equation}
\mathbi{q}_{k+1} = \Phi_{k+1|k}(\boldsymbol\omega)\mathbi{q}_{k}
\label{quat_kine}
\end{equation}

where $\boldsymbol\omega$ is the angular velocity and $\Phi_{k+4|k}$ is the transition matrix from time $k-1$ to $k$ such as:

\begin{equation}
 \Phi_{k+1|k}(\boldsymbol\omega) = \begin{pmatrix}  \cos(0.5||\boldsymbol\omega||\Delta T)I_3 - [\boldsymbol \psi\times] & \boldsymbol\psi \\ -\boldsymbol\psi^T &  \cos(0.5||\boldsymbol\omega||\Delta T)  \end{pmatrix} 
\end{equation}

\begin{equation}
 \boldsymbol \psi = \sin(0.5||\boldsymbol\omega||\Delta T)\frac{\boldsymbol\omega}{||\boldsymbol\omega||}
\end{equation}

@@

\begin{equation}
[\textbf{v} \times] = \begin{pmatrix} 0 & v_z & -v_y \\ -v_z & 0 & v_x \\ v_y & -v_x & 0 \end{pmatrix} 
\label{skewsymmat}
\end{equation}
Then, the attitude matrix $R \in SO(3) $ related to the quaternion $\mathbi{q}$ is given by :

\begin{equation}
R(\mathbi{q}) = (q_w^2 - |\textbf{q}|^2)I + 2 \textbf{q}\textbf{q}^T - 2q_w[\textbf{q} \times]
\label{quat_to_rot}
\end{equation}

Where $SO(3)$ is the special orthogonal group defined by:

\begin{center}
$SO(3) = \{ R \in \mathbb{R}^{ 3 \times 3} | RR^T = R^TR = I, det(R) = 1 \}$
\end{center}

@@

\subsubsection{Kalman Filtering Methods}

Introduced in 1960 by Kalman \cite{kalman_new_1960}, the Kalman filter is an algorithm that uses a series of measurements observed over time, containing noise, in order to estimate unknown variables (state vector). %The goal is to obtain more precise results than those based on a single measurement alone. 
Based on a model it is possible to obtain a more robust estimation of the state vector. Consider the following linear system, described by the difference equation and the observation model with additive noise:

\begin{equation}
\left\{ \begin{array}{l}
\textbf{x}_{k} = F_{k}\textbf{x}_{k-1}+B_{k}\textbf{u}_{k}+G_{k}\textbf{v}_{k}\\
\textbf{z}_{k} = H_{k}\textbf{x}_{k}+D_{k}\textbf{w}_{k}
\end{array} \right.
\label{linear_system}
\end{equation}

@@

The figure \ref{State_equation_discrete_case} represents the dynamical system described in \ref{linear_system} . A brief presentation of the algorithm is provided here,  see \cite{terejanu2013discrete} for more details. The algorithm has 3 important steps :\\

\begin{itemize}
\item The first step is the \textbf{prediction} of state $\textbf{x}^-_{k}$ at time $k$ and covariance $P^-_{k}$ is given by :

\begin{equation}
\left\{ \begin{array}{cl}
\textbf{x}^-_{k} = & F_{k}\textbf{x}^+_{k-1} + B_{k}\textbf{u}_{k} \\
P^-_{k} = & F_{k}P^+_{k-1}F^T_{k}+G_{k}Q_{k}G^T_{k}
\end{array}
\right.
\end{equation}

\item The \textbf{innovation} is the difference between predicted measurement and real measurement. It measure up the deviation used by the Kalman gain.
\begin{equation}
\left\{ \begin{array}{cl}
\boldsymbol\nu_{k} = & \textbf{z}_{k}-H_{k}\textbf{x}^-_{k} \\
S_{k} = & R_{k} +H_{k}P^-_{k}H^T_{k}
\end{array}
\right.
\end{equation}

\item And finally, An \textbf{update} is processed to correct the estimate state $\textbf{x}^+_{k} $ and his covariance $P^+_{k}$ with the innovation.

\begin{equation}
\left\{ \begin{array}{cl}
\textbf{x}^+_{k} = & \textbf{x}^-_{k} + K_{k}\boldsymbol\nu_{k} \\
P^+_{k} = &P^-_{k} - K_{k}S_{k}K^T_{k}
\end{array}
\right.
\end{equation}

@@

Where $K_k$ is the Kalman gain  representing relative importance of innovation  $\nu_{k}$.

\begin{equation}
K_{k}=P^-_{k}H^T_{k}S^{-1}
\end{equation}

%And so on for future state $k+1,...$.

@@

First, we propagate the estimate attitude and the gyro bias with equation \ref{prediction_cgo}.

\begin{equation}
\left\{\begin{array}{l}
\mathbi{q}_{k}^- = \Phi_{k|k-1}(\boldsymbol\omega_{gyro} - \boldsymbol \beta_{k-1}^+)\mathbi{q}_{k-1}^+ \\
\boldsymbol \beta_k^- = \boldsymbol \beta_{k-1}^+
 \end{array}
\right.
\label{prediction_cgo}
\end{equation}

\begin{itemize}
\item Correction
\end{itemize}

The idea is to compute a  constraint  given by a comparison between  an estimate value of a sensor and his real data at a time $k$.

\begin{equation}
\left\{\begin{array}{l}
\textbf{v}_{acc}^- = R_(\mathbi{q}_{k}^-)\textbf{a}^a + \boldsymbol  \beta_{acc} \\
\textbf{v}_{mag}^- = R_(\mathbi{q}_{k}^-)\textbf{m}^a + \boldsymbol  \beta_{mag}
 \end{array}
\right.
\end{equation}

All these vectors are normalized.

The '' gain '' is given by,

\begin{equation}
\boldsymbol \omega_{mes} = -vex(\sum_{i=1}^n \frac{k_i}{2}(\textbf{v}_i(\textbf{v}_i^-)^T -  \textbf{v}_i^-(\textbf{v}_i)^T )) 
\label{cgo_gain}
\end{equation}

where the $vex$ operator define by the inverse of the cross product.
\begin{equation}
\left\{\begin{array}{l}
\textbf{a}(\textbf{b})^T -  \textbf{b}(\textbf{a})^T  = [ (\textbf{a} \times \textbf{b}) \times ]\\
vex([\textbf{a}\times])  = \textbf{a}
 \end{array}
\right.
\end{equation}

Then, we the gain is used to correct the predicted attitude $\mathbi{q}_{k}^-$ by the following equation:

\begin{equation}
\left\{\begin{array}{l}
\mathbi{q}_{k}^+ = \mathbi{q}_{k}^- + \frac{\Delta T}{2} \Xi(\mathbi{q}_{k}^-) \boldsymbol\omega_{mes}k_p\\
\boldsymbol \beta_k^+ = \boldsymbol \beta_{k}^-  - \boldsymbol\omega_{mes}k_i
 \end{array}
\right.
\end{equation}

@@

Given the two sets of $n$ vectors $\{ \textbf{r}_{1},...,\textbf{r}_{n} \}$ and $\{ \textbf{b}_{1},...,\textbf{b}_{n} \}$, $n \geqslant 2 $, where each pair $(\textbf{r}_{i},\textbf{b}_{i})$ corresponds to a generalised vector, $\textbf{x}_{i}$, find the proper orthogonal matrix, $A$, which bring the first into the best least squares coincidence with the second. That is, find $A$ which minimises the function $J$ defined as:

@@

subject to the constraint $A^TA = I$ and $det(A)=1$. The equation (\ref{Wahba_loss_function}) can be written with a quaternion representation which gives:

@@

The minimisation problem from Equation (\ref{QUEST_bis}) is now a maximisation problem. thus, the solution $\mathbi{q}_{opt}$ is the eingen vector correspond to the largest eigen value of $K$ \cite{markley1999estimate}. The main problem of QUEST is that it does not take into account of all the previous observations performed since the beginning. A recursive QUEST algorithm (REQUEST) proposed by Bar-Itzhack manages this problem. According to the REQUEST algorithm, the propagation of $K_k$ is given by

\begin{equation}
K_k^- = \Phi_{k|k-1}(\boldsymbol\omega_{gyro})K_{k-1}^+\Phi_{k|k-1}^T(\boldsymbol\omega_{gyro})
\end{equation}

We compute $\delta K_k $ as

\begin{equation}
\delta K_k = \frac{1}{\sum_ia_i}\begin{pmatrix} \delta S_k-\delta\sigma_kI_3 & \delta\textbf{z}_k \\ \delta\textbf{z}_k^T & \delta\sigma_k
\end{pmatrix}
\end{equation}

\begin{equation}
\begin{array}{cc}
\delta B_k = \sum_i a_i \textbf{b}_i\textbf{r}_i^T & \delta S_k = B_k + B_k^T\\
\delta \textbf{z}_k = \sum_ia_i\textbf{b}_i\times\textbf{r}_i & \delta \sigma_k = tr(B_k)
\end{array}
\end{equation}

Where $\textbf{b}_i$ is the observation vector at time $k$. Then we update the K-matrix.

\begin{equation}
K_k^+ = \frac{\rho_k m_{k-1}}{m_k}K_k^- + \frac{1}{m_k}\delta K_k
\end{equation}

A drawback is that the value of $\rho_k$ comes from heuristics. Choukroun revisit REQUEST provides an improvement with his Optimal REQUEST \cite{choukroun_novel_2003} where an optimal $\rho_k^*$ is computed which gives better results. Here we choose to implement this version of REQUEST. Contrary to other methods, the QUEST algorithms do not estimate the gyroscope bias. Which can make it less robust in some cases.
There are also other methods based on Wahba's problem like the Davenport's q-method \cite{weighted1971nasa} proposed in 1971 considered more robust than QUEST. Or the method ESOQ proposed by Mortari \cite{mortari1997esoq}  in 1997. But in this study we limit our investigation to REQUEST.\\

@@

The first step is to propagate all particles $\{\textbf{x}^{(i)}_{k-1},w^{(i)}_{k-1}\}$ to $\{\textbf{x}^{(i)}_{k},w^{(i)}_{k}\}$ with the nonlinear function $f$ while the weight remain unchanged :

\begin{equation}
\textbf{x}^{(i)}_{k} = f(\textbf{x}^{(i)}_{k-1},\textbf{u}_{k-1}) + \textbf{w}_{k-1}^{(i)}\\
\end{equation}

where $N$ sample $\textbf{w}_{k-1}^{(i)}$ of the process noise (supposed gaussian in our case) are drawn.

@@

At the update step, the weights associated with each particle is updated based on the likelihood function $p( \textbf{z}_k| \textbf{x}_k^{(i)})$.
\begin{equation}
\left\{ 
\begin{array}{l}
w_k^{(i)} = w_{k-1}^{(i)} p( \textbf{z}_k| \textbf{x}_k^{(i)})\\
\tilde{w}_k^{(i)} = \frac{w_{k}^{(i)}}{\sum_iw_{k}^{(i)}}
\end{array}
\right.
\end{equation}

The weight are normalised such as we have $\sum_i\tilde{w}_k^{(i)}=1$. To finish, the mean is computed with the following equation:

\begin{equation}
 \textbf{x}_k^+ = \sum_{i=1}^N\tilde{w}_k^{(i)} \textbf{x}_k^{(i)}
 \end{equation}

@@

The choice of the threshold is important. Indeed, a high threshold gives $N_{eff} \approx N$, but that does not mean necessarily we have a lot of efficient particles. On the contrary, it means that most of the particle are identical making large computations unnecessary. Moreover, with a low threshold, many particles are neglected making computations less efficient. In general, the threshold is fixed at  $\frac{N}{2}$ or $\frac{N}{3}$.\\

There are several types of resampling, a comparison is made in \cite{douc2005comparison}. We took the resampling algorithm described in \cite{arulampalam2002tutorial} because it is the most commun. Cheng, Yang and Crassidis have propose an attitute estimation with a particle filter method \cite{cheng_particle_2010}. We have also considered the revisited algorithm of Cheng \cite{chang_particle_2014} based on bootstrap filter\cite{gordon1993novel}. We have chosen to implement this one because they use the same approach as USQUE \cite{crassidis_unscented_2003} and MEKF\cite{markley2003attitude}. We decided to explore the performance of this three algorithms with the same model. \\

The particle filter provides an estimate of the nonlinear system without assuming that the noise is Gaussian. So it is more robust to any situation, which makes it more performant in the context of a real applications. But an important computation time is required and that is a limitation for a real-time application.

@@

We have access through our designed ground truth to the true attitude $\mathbi{q}^{real}_k$ at time $k$ of the reference frame $\{c\}$ related to $\{a\}$.
By computation, each method returns an estimate attitude $\mathbi{q}^{est}_k$ at time $k$ (see figure \ref{test_method}).

@@

@@

>>>>

\subsection{Quaternion representation}

@@

For attitude representation, the quaternion must satisfies a single constraint given by  $|| \mathbi{q}  || = 1$.

Then, the attitude matrix $R \in SO(3) $ related to the quaternion $\mathbi{q}$ is given by :
\vspace{-0.15cm}
\begin{equation}
R(\mathbi{q}) = (q_w^2 - |\textbf{q}|^2)I + 2 \textbf{q}\textbf{q}^T - 2q_w[\textbf{q} \times]
\label{quat_to_rot}
\end{equation}

Where $SO(3)$ is the special orthogonal group defined by:
\vspace{-0.15cm}
$$SO(3) = \{ R \in \mathbb{R}^{ 3 \times 3} | RR^T = R^TR = I, det(R) = 1 \}$$

@@

\begin{equation}
[\textbf{v} \times] = \begin{pmatrix} 0 & v_z & -v_y \\ -v_z & 0 & v_x \\ v_y & -v_x & 0 \end{pmatrix} 
\label{skewsymmat}
\end{equation}

@@

\subsubsection{Kalman Filter}

Introduced in 1960 by Kalman \cite{kalman_new_1960}, the Kalman filter is an algorithm that uses a series of measurements observed over time, containing noise, in order to estimate unknown variables (state vector). Based on a model it is possible to obtain a more robust estimation of the state vector. Consider the following linear system, described by the difference equation and the observation model with additive noise:

\begin{equation}
\left\{ \begin{array}{l}
\textbf{x}_{k+1} = F_{k}\textbf{x}_{k}+B_{k}\textbf{u}_{k}+G_{k}\textbf{v}_{k}\\
\textbf{z}_{k} = H_{k}\textbf{x}_{k}+D_{k}\textbf{w}_{k}
\end{array} \right.
\label{linear_system}
\end{equation}

@@

A brief presentation of the algorithm is provided here,  see \cite{terejanu2013discrete} for more details. The algorithm has 3 important steps :\\

\begin{itemize}
\item The first step is the \textbf{prediction} of state $\textbf{x}^-_{k+1}$ at time $k+1$ and covariance $P^-_{k+1}$ is given by :

\begin{equation}
\left\{ \begin{array}{cl}
\textbf{x}^-_{k+1} = & F_{k}\textbf{x}^+_{k} + B_{k}\textbf{u}_{k} \\
P^-_{k+1} = & F_{k}P^+_{k}F^T_{k}+G_{k}Q_{k}G^T_{k}
\end{array}
\right.
\end{equation}

\item The \textbf{innovation} is the difference between predicted measurement and real measurement. It measure up the deviation used by the Kalman gain.
\begin{equation}
\left\{ \begin{array}{cl}
\boldsymbol\nu_{k+1} = & \textbf{z}_{k+1}-H_{k+1}\textbf{x}^-_{k+1} \\
\mathcal{S} = & R_{k+1} +H_{k+1}P^-_{k+1}H^T_{k+1}
\end{array}
\right.
\end{equation}

\item And finally, An \textbf{update} is processed to correct the estimate state $\textbf{x}^+_{k+1} $ and his covariance $P^+_{k+1}$ with the innovation.

\begin{equation}
\left\{ \begin{array}{cl}
\textbf{x}^+_{k+1} = & \textbf{x}^-_{k+1} + \mathcal{K}_{k+1}\boldsymbol\nu_{k+1} \\
P^+_{k+1} = &P^-_{k+1} - \mathcal{K}_{k+1}S_{k+1}\mathcal{K}^T_{k+1}
\end{array}
\right.
\end{equation}

@@

Where $\mathcal{K}_{k+1}$ is the Kalman gain  representing relative importance of innovation  $\nu_{k+1}$.

\begin{equation}
\mathcal{K}_{k+1}=P^-_{k+1}H^T_{k+1}\mathcal{S}^{-1}
\end{equation}

The Kalman filter provides an optimal estimation for linear systems. Nevertheless, most of studied dynamical systems are nonlinear (attitude estimation, pendula, ... ). Thus, extended Kalman Filter approach  was proposed for nonlinear systems \cite{larson1967application, larson1967precomputation,terejanu2008extended}.

Here, we have chosen to use quaternions for the attitude representation because of its singular-free specificity. But special care must be taken when designing a quaternion based extended Kalman filter. Because of the constraint on the quaternions, the covariance matrix $P$ can be singular \cite{vik2009integrated} which is problematic for the numerical stability. There exist several methods to deal with this issue. Multiplicative Extended Kalman Filter (MEKF)\cite{markley2003attitude} and the Additive Extended Kalman Filter (AEKF) \cite{bar1991quaternion} are designed to tackle this problem.The difference between AEKF and MEKF is discussed in \cite{markley_multiplicative_2004}. For the library, we choose to implement MEKF.

On the other hand, the Unscented Kalman Filter method was proposed by Julier and Uhlmann \cite{julier_new_1995} in 1995 . It belongs to a class of filters called Sigma-Point Kalman Filters. Instead of using the Taylor series expension, UKF use an unscented transformation \cite{uhlmann1995dynamic} to approximate the nonlinear projection of mean and covariance of a probability distribution. For the library, we choose to implement the filter proposed by Crassidis et al. called UnScented QUaternion Estimator (USQUE) \cite{crassidis_unscented_2003}, based on UKF. They used the same representation as MEKF.

@@

@@

Given the two sets of $n$ vectors $\{ \textbf{r}_{1},...,\textbf{r}_{n} \}$ and $\{ \textbf{b}_{1},...,\textbf{b}_{n} \}$, $n \geqslant 2 $, where each pair $(\textbf{r}_{i},\textbf{b}_{i})$ corresponds to a generalised vector, $\textbf{x}_{i}$, find the proper orthogonal matrix, $A$, which bring the first into the best least squares coincidence with the second. That is, find $A \in SO(3)$ which minimises the function $J$ defined as:

@@

The equation (\ref{Wahba_loss_function}) can be written with a quaternion representation which gives:

@@

The minimisation problem from Equation (\ref{QUEST_bis}) is now a maximisation problem. thus, the solution $\mathbi{q}_{opt}$ is the eingen vector correspond to the largest eigen value of $K$ \cite{markley1999estimate}. The main problem of QUEST is that it does not take into account of all the previous observations performed since the beginning. A recursive QUEST algorithm (REQUEST) proposed by Bar-Itzhack manages this problem.  Contrary to other methods, the QUEST algorithms do not estimate the gyroscope bias. Which can make it less robust in some cases.

There are also other methods based on Wahba's problem like the Davenport's q-method \cite{weighted1971nasa} proposed in 1971 considered more robust than QUEST. Or the method ESOQ proposed by Mortari \cite{mortari1997esoq}  in 1997. However, for this library, we limit our investigation to REQUEST.\\

@@

The first step is to propagate all particles $\{\textbf{x}^{(i)}_{k},w^{(i)}_{k}\}$ to $\{\textbf{x}^{(i)}_{k+1},w^{(i)}_{k+1}\}$ with the nonlinear function $f$ while the weight remain unchanged :

\begin{equation}
\textbf{x}^{(i)}_{k+1} = f(\textbf{x}^{(i)}_{k},\textbf{u}_{k}) + \textbf{w}_{k}^{(i)}\\
\end{equation}

where $N$ sample $\textbf{w}_{k}^{(i)}$ of the process noise (supposed gaussian in our case) are drawn.

@@

At the update step, the weights associated with each particle is updated based on the likelihood function $p( \textbf{z}_{k+1}| \textbf{x}_{k+1}^{(i)})$.
\begin{equation}
\left\{ 
\begin{array}{l}
w_{k+1}^{(i)} = w_{k}^{(i)} p( \textbf{z}_{k+1}| \textbf{x}_{k+1}^{(i)})\\
\tilde{w}_{k+1}^{(i)} = \frac{w_{k+1}^{(i)}}{\sum_iw_{k+1}^{(i)}}
\end{array}
\right.
\end{equation}

The weight are normalised such as we have $\sum_i\tilde{w}_{k+1}^{(i)}=1$. To finish, the mean is computed with the following equation:

\begin{equation}
 \textbf{x}_{k+1}^+ = \sum_{i=1}^N\tilde{w}_{k+1}^{(i)} \textbf{x}_{k+1}^{(i)}
 \end{equation}

@@

The choice of the threshold is important. Indeed, a high threshold gives $N_{eff} \approx N$, but that does not mean necessarily we have a lot of efficient particles. On the contrary, it means that most of the particle are identical making large computations unnecessary. Moreover, with a low threshold, many particles are neglected making computations less efficient. In general, the threshold is fixed at  $\frac{N}{2}$ or $\frac{N}{3}$. There are several types of resampling, a comparison is made in \cite{douc2005comparison}. We took the resampling algorithm described in \cite{arulampalam2002tutorial} because it is the most commun. Cheng, Yang and Crassidis have propose an attitute estimation with a particle filter method \cite{cheng_particle_2010}. We have also considered the revisited algorithm of Cheng \cite{chang_particle_2014} based on bootstrap filter\cite{gordon1993novel}. We have chosen to implement this one because they use the same approach as USQUE \cite{crassidis_unscented_2003} and MEKF\cite{markley2003attitude}. We decided to explore the performance of this three algorithms with the same model. The particle filter provides an estimate of the nonlinear system without assuming that the noise is Gaussian. So it is more robust to any situation, which makes it more performant in the context of a real applications. But an important computation time is required and that is a limitation for a real-time application.

@@

\subsection{Score definition}
We have access through our designed ground truth (simulated or from computer vision technics) to the true attitude $\mathbi{q}^{real}_k$ at time $k$ of the reference frame $\{c\}$ related to $\{a\}$.
By computation, each method returns an estimate attitude $\mathbi{q}^{est}_k$ at time $k$ (see figure \ref{test_method}).

@@

\subsection{Simulated environment}

Our framework has been entierely coded in C++ and all tests have been launched on a Dell Optiplex 9020 with Intel $ Core^{TM}$ i7-4790 CPU at 3.60GHz $\times$ 8.  All curves were generated with \texttt{R} and \texttt{Matlab}. No paralelization or code optimisation have been done, our program turns entirely in sequential manner. Our goal is to see the difference of computation duration of all methods without modifying the performance with a specific optimisation.

\begin{table}
\begin{tabular}{|c|c|c|}
\hline
Sensor & Bias instability  & Measurement noise\rule[-2pt]{0pt}{10pt} \\
\hline
\hline
 Gyroscope & $0.316\times 10^{-2} $  $\mu$rad.s$^{-\frac{1}{2}}$ & $ 3.16$ $ \times 10^{-4} $ $\mu$rad.s$^{-\frac{3}{2}}$ \rule[-1.5pt]{0pt}{13pt}\\
 \hline
Magnetometer & $0.0023$ mGauss$/\sqrt{Hz}$ & 0.05 mGauss  \rule[-1.5pt]{0pt}{13pt}\\
 \hline
Accelerometer &  $0.002$ m.s$^{-2}/\sqrt{Hz} $ & 0.02 m.s$^{-2}$ \rule[-1.5pt]{0pt}{13pt}\\
 \hline
\end{tabular}
\caption{Calibrated IMU data performance specification, Gyrometer, Accelerometer, Magnetometer}
\label{spec_imu}
\end{table}

For our Sensor Model, we've used  specification described in table \ref{spec_imu}.

@@

\subsection{Experimental validation}

We have designed a system that can be easily repeatable in laboratory. We used a smartphone as our IMU and a calibrated webcam (see figure \ref{situation_validation} ). With this system, we have defined our ground-truth as follows. A marker is stuck on the smartphone and we calculate an estimate of the attitude of the marker (which represents the smartphone) relative to the reference frame of the webcam $\{w\}$. We used the library \texttt{ArUco}  as optical feedback. At the same time, we get the data from the IMU of the smartphone to proceed the attitude estimation with our different algorithms.

\begin{figure}
\centering
\includegraphics[scale=0.35]{images/situation_validation.png}
\caption{Representation of different reference frame used in this experiment. Here, only $ \text{\mathbi{q}}_w^a $  does not change.}
\label{situation_validation}
\end{figure}

At time $t = 0$, the camera and the smartphone has been set such as attitude quaternions $\mathbi{q}^a_w$, $\mathbi{q}^c_a$ et $\mathbi{q}^w_c$ are approximatively known. The camera being fixed, $\mathbi{q}^a_w $ does not vary. Moreover, from figure \ref{situation_validation}  we have:

\begin{equation}
\mathbi{q}^c_a =( \mathbi{q}^c_w \otimes \mathbi{q}^a_w )^{-1}
\end{equation}

\begin{figure}
\centering
\includegraphics[scale=0.35]{images/output_aruco.png}
\caption{exemple of output of \texttt{ArUco} with 6 markers. Here the object-fixed reference frame $\{c\}$ is drawn}
\label{situation_validation}
\end{figure}

All methods give an estimate of $\mathbi{q}^c_a$ , that are compared with the data from the camera. The optical feedback provides a ground truth estimate for this experiment. To estimate the error that can occur with the optical feedback. We were careful in the experimentation to avoid the occlusion of the marker that is required to estimate the orientation. During this experiment, we applied a rotation of 90 degree around the $z_c$  on the smartphone in clockwise direction. Then, another rotation of 90 degree around the $z_c$ on the smartphone but in anticlockwise direction to return to the initial state. In order to study possible divergence of all methods.
